Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-09 WholePlotTrajectories.tex
dicussion

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness, alone, gives limited or misleading information on forests
recovery and functioning \citep{Martin2015, Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness
that would reveal underlying ecological processes and (ii) composition
that is crucial for conservation issues
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes would
directly link communities diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect, the functional trait-based approach that focus on major traits
related to species ecology and mediate species performance in a given
environment was sucessfully adopted \citep{Diaz2005, Villeger2008a}. For
instance, the functional approach revealed in tropical rainforests the
deterministic processes entailing, after disturbance, a functional shift
from a dominance of ``conservative'' slow-growing species dealing with
scarce resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift is translated
into the trajectories of key functional traits related to resource
acquisition (leaf and stem traits) and life-history traits (seed mass,
maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of communities response to disturbance would
encompass the changes in functional redundancy, that quantifies the
amount of shared trait values among species \citep{Carmona2016}. The
high functional redundancy of hyperdiverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

@@

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the functional richness and evenness, \emph{i.e} Rao
index of quadratic entropy which combines species abundance distribution
and average pairwise dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analysed through polynomial regression between (i) taxonomic and
functional richness, evenness and diversity and (ii) the initial \%AGB
loss at 10, 20 and 30 years after disturbance.

@@

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
richness and evenness trajectories depended on the disturbance intensity
with their maximum positively correlated to \%AGB loss
\(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to maximum reached after 20-25 years and, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the inital values.

The second-degree polynomial regressions between (i) the \%AGB loss and
(ii) taxonomic and functional richness and evenness after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. Along the
disturbance intensity gradient the disturbance impact was more markedly
hump-shaped for the taxonomic richness than for the taxonomic evenness
and the functional diversity, the latest being almost linear. The
regression model better predicted the functional richness and evenness
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Evenness}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Evenness}<0.43\) respectively)

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
marked and consistent trajectories over post-disturbance time. in
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.87\) and
\(\rho_{spearman}^{functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

Communities taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years. The taxonomic differences
among communities, marked before disturbance by the distinct starting
points on the NMDS axis 2, were maintained throughout recovery
trajectories. More than commonly thought, post-disturbance trajectories
depended on community initial composition, that partly determined the
pool of recruited species and constrained the trajectories towards the
initial composition. The high resilience of communities taxonomy
revealed that species not belonging to the pre-disturbance community
were hardly recruited because of the commonness of dispersal limitation
among tropical tree species \citep{Svenning2005}.

Trajectories of community functional composition appeared similar in
time and close in terms of functional composition. Because
pre-disturbance surviving trees mirror the initial community
\citep{Herault2018}, changes in functional composition relied upon the
recruitment of species or functional types infrequent or absent before
disturbance. Competitive pioneers became dominant in filling the
environmental niches of high availability of light, space and nutrient
made available by the disturbance. It changed community functional
composition in the same way for all communities, towards more
resource-acquisitive strategies
\citep{Grime1998, Westoby1998, Wright2004, Reich2014}. Later during the
community recovery long-lived, more resistant and shade-tolerant
species, would exclude those pioneers and restore the pre-disturbance
composition.

@@

Disturbance appeared to enhance community taxonomic richness until an
intensity threshold (20-25\% of initial AGB loss), while the taxonomic
evenness was somewhat decoupled from the disturbance intensity as
already observed in the Guiana Shield \citep{Baraloto2012a} and in
Bornean tropical forests \citep{Cannon1998}. The IDH well predicted the
disturbance impact on community taxonomic richness and to some extent on
taxonomic evenness. The disturbance intensity determines the balance in
the community between pre-disturbance surviving trees and the trees
recruited afterward. In the Guiana Shield, the pool of true pioneer
species specifically recruited after disturbance is restricted to a few
common genera (e.g.~Cecropia spp., Vismia spp.). Below the intensity
threshold the size of the surviving community maintains its
pre-disturbance taxonomic richness and the recruitment of pioneers,
infrequent or absent before disturbance, increases both community
taxonomic richness and evenness. Beyond the intensity threshold the
taxonomic diversity of the surviving trees decreases and is not offset
by the taxonomic enrichment of pioneers. The disturbance impact was
similar but slighter for community taxonomic evenness, that is less
sensitive to the loss of rare species and rather represents the
increased dominance of pionneers that balanced the usual hyper-dominance
of tropical forest. While below the intensity threshold pioneers species
increase the diversity of dominant or common species, beyond the
threshold they become in turn highly dominant.

Conversely the IDH was disproved to predict the dsiturbance impact on
community functional diversity. Pioneer species were functionally highly
different from the pre-disturbance community composition, so their
recruitment increased the community functional richness and eveness
whenever its intensity.

Along time, the post-disturbance taxonomic richness trajectories showed
first a similar drop following the species loss during disturbance, and
then a species gain determined by the disturbance intensity. Until an
intensity threshold the gain in taxonomic richness increased with the
disturbance intensity through the establishment of long-lived pioneers
enhancing community taxonomic richness and evenness in the long term.
These long-lived pionneers were functionally quite different from the
functional composition and their establishment entailed a progressive
and long-lasting increase of the functional diversity. Beyond the
intensity threshold a few short-lived pioneers occupied the vacated
environmental space and prevented further species establishment. These
short-lived pioneers were functionally very different from the
pre-disturbance community and rapidly and significantly increased the
functional diversity. Already after 10 years, though, they started to
decline and the functional diversity started to decrease. Likely the
establishment of long-lasting pioneers will follow and we expect the
taxonomic and functional trajectories to catch up with those observed
after intermediate disturbance.

\subsection{The functional redundancy, key of the community
resilience}\label{the-functional-redundancy-key-of-the-community-resilience}

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. At first place
the pioneers recruited, functionally different from the pre-disturbance
functional composition did not compensate for the reundancy decrease.
Progressively though, first establish species were replaced by more
competitive long-lived pionneers or late-successional species resembling
more the pre-disturbance functional composition and restoring the
functional redundancy. This replacement was stochastic and followed the
lottery recruitment rules, that ease the recovery for the first
recruited species but then increasingly hamper it with the emergence of
interspecific competition due to increasing functional redundancy
\citep{Busing2002}. Along time the recovery of infrequent species is
increasingly difficult, and the recovery of the functional redundancy
that was just initiated after 30 years in some communities is difficult
to estimate \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of pre-disturbance communities and higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the long-term
recovery of infrequent species increases the risks to loose cornerstone
species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species might
indeed have unique functions in the ecosystem or be a key for some
fauna.

@@

{[}Our study revealed communities cyclic recovery after disturbance
allowing the resilience of their functioning and taxonomic composition
with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for 20 years after
disturbance through the enrichment of the communities with pioneers and
light-demanding species,in accordance with the IDH. The IDH, though,
poorly predicted the disturbance impact on communities taxonomic
richness and evenness that were blurried by the emergence of competitive
exclusion along time.{]} The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.

>>>>

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, though, gives limited or misleading information on
forests recovery and functioning \citep{Martin2015, Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness
that would reveal underlying ecological processes and (ii) composition
that is crucial for conservation issues
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes would
directly link communities diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect, the functional trait-based approach that focus on major traits
related to species ecology and mediate species performance in a given
environment was sucessfully adopted \citep{Diaz2005, Villeger2008a}. For
instance, the functional approach revealed in tropical rainforests the
deterministic processes entailing, after disturbance, a functional shift
from a dominance of ``conservative'' slow-growing species dealing with
scarce resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift is translated
into the trajectories of key functional traits related to resource
acquisition (leaf and stem traits) and life-history traits (seed mass,
maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of communities response to disturbance would
encompass the changes in functional redundancy, that quantifies the
amount of shared trait values among species \citep{Carmona2016}. The
high functional redundancy of hyperdiverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

@@

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the functional richness and functional evenness,
\emph{i.e} Rao index of quadratic entropy which combines species
abundance distribution and average pairwise dissimilarity based on
species functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analysed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

@@

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
functional richness and evenness trajectories depended on the
disturbance intensity with their maximum positively correlated to \%AGB
loss \(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to a maximum reached after 20-25 years, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the inital values.

The second-degree polynomial regressions between (i) the \%AGB loss and
(ii) taxonomic and functional richness and evenness after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the taxonomic
evenness. For both functional richness and evenness the relationship was
almost linear. The regression model better predicted the functional
richness and evenness (\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Evenness}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Evenness}<0.43\) respectively)

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
marked and consistent trajectories over post-disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.87\) and
\(\rho_{spearman}^{functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

Communities taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years.

The taxonomic differences among communities, marked before disturbance
by the distinct starting points on the NMDS axis 2, were maintained
throughout recovery trajectories. More than commonly thought,
post-disturbance trajectories depended on community initial composition,
that partly determined the pool of recruited species and constrained the
trajectories towards the initial composition. The high resilience of
communities taxonomy revealed that species not belonging to the
pre-disturbance community were hardly recruited because of the
commonness of dispersal limitation among tropical tree species
\citep{Svenning2005}.

Conversely, disturbed communities followed functional trajecotories
similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrient vacated by the
disturbance. The recruitment of pioneers similarly changed community
functional composition towards more resource-acquisitive strategies
\citep{Grime1998, Westoby1998, Wright2004, Reich2014}. Thereafter, again
in the same way for all disturbed communities, long-lived, more
resistant and shade-tolerant species excluded the first established
pioneers and started the recovery of pre-disturbance functional
composition.

@@

The IDH predicted well the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.). Below the intensity
threshold the size of the surviving community maintained the
pre-disturbance taxonomic richness while the recruitment of pioneers,
infrequent or absent before disturbance, increased both community
taxonomic richness and evenness. Beyond the intensity threshold, the
disturbance decreased the taxonomic richness of surviving trees which
was not offset by the enrichment of pioneers, so that the overall
community taxonomic richness decreased according to the disturbance
intensity. For community taxonomic evenness the disturbance impact was
similar but slighter, as the evenness is less sensitive to the loss of
rare species. Taxonomic evenness rather represented the increasing
dominance of pionneers that balanced the usual hyper-dominance of
tropical forest below the intensity threshold, thus increasing community
overall evenness up to the intensity threshold beyond which pioneers
became in turn highly dominant and decreased the overall evenness.

Conversely the IDH was disproved regarding the disturbance impact on
community functional diversity. Irrespective of the disturbance
intensity the recruitment of pionneers, functionally highly different
from the composition of pre-disturbance community, increased the
community functional richness and eveness.

Along time, taxonomic richness trajectories for all disturbed
communities first dropped similarly following the loss of species during
disturbance. Then, trajectories showed a species gain depending on the
disturbance intensity. The species gain was all the more significant as
the disturbance intensity increased, up to an intensity threshold,
through the establishment of long-lived pioneers enhancing community
taxonomic richness and evenness in the long term. These long-lived
pionneers, functionally quite different from the functional composition,
entailed in parallel a progressive and long-lasting increase of the
functional diversity. Beyond an intensity threshold, though, a few
short-lived pioneers occupied the vacated environmental space and
prevented the establishment of other species. These short-lived pioneers
were functionally very different from the pre-disturbance community and
entailed a rapid and significant increase of functional richness en
evenness. Already after 10 years, though, short-lived pioneers started
to decline and the functional richness and evenness decreased. Likely
the establishment of long-lasting pioneers will follow this decrease and
we expect the taxonomic and functional trajectories to catch up with
those observed after intermediate disturbance.

\subsection{The functional redundancy, key of community
resilience}\label{the-functional-redundancy-key-of-community-resilience}

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the pionneers first
recruited were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was difficult to estimate
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of the pre-disturbance communities, with higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the long-term
recovery of infrequent species increases the risks to loose cornerstone
species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species might
indeed have unique functions in the ecosystem or be a key for some
fauna.

@@

Our study revealed the post-disturbance recovery of both taxonomic
composition, maintaining the initial differences among communities, and
functional composition, that was in contrast similar among communities.
The IDH was validated for community taxonomic richness and, to some
extent, taxonomic evenness but disproved regarding community functional
richness and evennes that were enhanced for any disturbance intensity by
the high functional differences of pioneers compared to
late-successional functoinal composition. The IDH was translated in time
by the recruitment, beyond an intensity threshold, of short-lived
pionneers that prevented in the first times the establishment of more
diverse long-lived pionneers, recruited otherwise below the intensity
threshold. The resilience of tropical forests proved consistent although
several decades-long. Still, the disturbance impact on communities
redundancy cautioned against the risks of infrequent species loss and
the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.
